---
title: "DESeq2"
output: html_document
date: "2023-07-25"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Gene-level differential expression analysis using DESeq2

# load libraries
```{r}
library(tidyverse)
library(RColorBrewer)
library(DESeq2)
library(pheatmap)
library(dplyr)
library(rlang)
library(stringr)
library(ggplot2)
library(ggrepel)
library(RColorBrewer)
```

#load data

Here the count tables are loaded along with the corresonding metadata of the experiment. Since the counttables here contain numbered genes rather than gene names, a separate file is given with correct translations to cyanobase IDs

```{r}
## read count data
txt_files_ls = paste0("data/",list.files(path="data/", pattern="*.txt"))
txt_files_df <- lapply(txt_files_ls, function(x) {
  read.table(file = x, 
             header = F, 
             sep ="\t",
             row.names = 1, 
             col.names = c("genes", paste0("S",str_extract({{x}}, "\\d+"))))
  })
countData <- do.call("cbind", txt_files_df)
countData <- filter(countData, !str_detect(rownames(countData),"_")) %>% 
  filter(rowSums(.)>24)

## translate gene IDs
conv <- read.csv("meta/translate_genes.csv", row.names = 2)
conv['gene2045',]$Final_name = 'sll0944a'
countData <- merge(countData,conv['Final_name'],by='row.names') #genes decrease due to some gene numbers being rRNA having no biocyc ID
rownames(countData) <- countData$Final_name
countData <- countData[,!names(countData)%in%c("Final_name","Row.names")]
KO <- c("sll0404","slr0806","slr1556","sll0587","slr0721","slr0301","sll0107")
countData <- countData[!rownames(countData) %in% KO,] #remove knocked out genes

## load metadata
meta <- read.table("meta/meta.txt", header=T)

```


#quality control

first we plot the count distribution, it is expected to observe a gamma distribution.

```{r}
ggplot(countData) +
  geom_histogram(aes(x=S01),stat = "bin", bins = 200) +
  xlim(-5, 5000)  +
  xlab("Raw expression counts") +
  ylab("Number of genes")
```

next we look at the mean vs the variance, it is expected to see the variance increase with respect to the mean as the mean increases.

```{r}
mean_counts <- apply(countData, 1, mean)
variance_counts <- apply(countData, 1, var)
df <- data.frame(mean_counts, variance_counts)

ggplot(df) +
        geom_point(aes(x=mean_counts, y=variance_counts)) + 
        geom_line(aes(x=mean_counts, y=mean_counts, color="red")) +
        scale_y_log10() +
        scale_x_log10()
```

make a DESeq2 matrix which can be normalised and used to observe variation (PCA) and correlation (heatmap) between samples 

```{r}
dds <- DESeqDataSetFromMatrix(countData=countData, 
                              colData=meta, 
                              design=~strain)
```

```{r}
#normalise with respect to library size and apply log2 scale
rld <- rlog(dds, blind=TRUE) 
	
# Plot PCA 
plotPCA(rld, intgroup="strain") + 
  geom_label(aes(label = rld@colData@rownames)) + 
  xlim(-10,12) +
  ylim(-5,5)
 
# plot heatmap
rld_mat <- assay(rld) #Extract the rlog matrix from the object
rld_cor <- cor(rld_mat)  #Compute pairwise correlation values
pheatmap(rld_cor)

```

## differential expression analysis

create deseq2 object and extract the normalized count table and fold changes
```{r}
dds <- DESeq(dds)
normalized <- as.data.frame(counts(dds, normalized=TRUE))
res <- results(dds)
head(results(dds, tidy=TRUE))
summary(res)

## Turn the results object into a data frame
res_df <- data.frame(res)

## Subset the significant results
sig_res <- filter(res_df, padj < 0.05)

```

plot dispersion
```{r}
plotDispEsts(dds)
```

plot a vulconaplot

```{r}
My_Theme = theme(
  axis.title.x = element_text(size = 15),
  axis.text.x = element_text(size = 16),
  axis.title.y = element_text(size = 15),
  )

x <- res_df
x$lab <- ""
x$sign <- NA
x$lab[abs(x$log2FoldChange) >= 1 & x$padj<0.05] <- "sig_large"
x$sign[abs(x$log2FoldChange) >= 1 & x$padj<0.05] <- rownames(x[abs(x$log2FoldChange) >= 2 & x$padj<0.05,])
x$lab[abs(x$log2FoldChange) < 1 & x$padj<0.05] <- "sig_small"
x$lab[x$padj>=0.05] <- "not_sig"

ggplot(data=x, aes(x=log2FoldChange, y=-log10(pvalue),col=lab,label=x$sign)) + geom_point() + scale_color_manual(values=c("black","red","blue")) + My_Theme
```


test for enriched gene sets
```{r}
library(clusterProfiler)
gene <- res_df[,"log2FoldChange"]
names(gene) <- row.names(res_df)
gene<-na.omit(gene)
gene <- sort(gene, decreasing = TRUE)

kk <- gseKEGG(geneList     = gene,
               organism     = 'syn',
               minGSSize    = 120,
               pvalueCutoff = .1,
               verbose      = FALSE)

dotplot(kk)

```



